---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r settings, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = TRUE,
  echo = TRUE,
  include = TRUE
)
```

# regridR: re-grid raster layers

Most global-coverage raster layers containing variables used in species distribution models come in unprojected longitude-latitude coordinates, with a nominal spatial resolution like "approximately 1x1 km2 at the equator". One problem is that lon-lat pixels are not equal-area, and they're not really 1x1-km2 either (or whatever is their nominal resolution): Because the longitude meridians converge towards the poles, lon-lat pixels cover progressively smaller areas as we move away from the Equator, and they are already considerably smaller than their nominal resolution across temperate areas like Europe. Their actual sizes can be checked e.g. with the `terra::cellSize()` function.

If we want our raster variables on a grid of pixels that matches an equal-area grid within a given region (such as the EEA reference grid in Europe), we can use the `regrid()` function to convert raster layers into such equal-area grid. Below is a worked example. We start by downloading some variable layers from the CHELSA website, using the `downloadif()` function also included in the `regridR` package. This will download files if they haven't already been completely downloaded and saved in the destination folder:

```{r, results='hide'}
# LOAD REQUIRED PACKAGES ----

library(regridR)

library(terra)


# DOWNLOAD SOME VARIABLES ----

# get CHELSA climate links for a couple variables:
links <- linkbuild(c("bio1", "scd"))

# create a folder for receiving downloads:
dir.create("outputs/variables", recursive = TRUE)

# allow longer download times:
options(timeout = 6000)

# download variables if not already there:
downloadif(links, destdir = "outputs/variables")

# import variables from downloads folder:
layers <- terra::rast(list.files("outputs/variables", full.names = TRUE))

terra::plot(layers, nc = 1)
```

Next, we will import a vector polygon map of an equal-area grid recommended by the European Environment Agency (EEA):

```{r}
# IMPORT EQUAL-AREA VECTOR GRID ----

EEAgrid <- terra::vect(system.file("extdata/eea10_belgium.gpkg", 
                                   package = "regridR"))

terra::plot(EEAgrid)
```

We can project the vector grid to overlay a part of the climate layers, and see that they don't align (and the pixels are not equal-area), so simply aggregating the raster pixels wouldn't be a good option:

```{r}
terra::plot(layers[[1]], maxcell = ncell(layers), ext = c(5, 5.4, 49.5, 49.7))

terra::plot(terra::project(EEAgrid, layers), add = TRUE)
```

So, we'll use the `regrid()` function of the `regridR` package to get the climate layers on a raster grid whose pixels match the input EEA grid cells:

```{r, eval=FALSE}
# RE-GRID LAYERS ----

layers_regrid <- regridR::regrid(layers = layers, grid = EEAgrid, 
                                 na.rm = TRUE, touches = TRUE)

terra::plot(layers_regrid)
```

By default, `regrid()` will use the `mean()` function to summarize the values of the pixels falling within each polygon grid cell, and the `terra::zonal()` function to do this summarizing. However, installing also the `exactextractr` package and **running `regrid()` with the argument `exactextract = TRUE`** can make the computation **considerably faster** for large grids, albeit with a slightly different algorithm:

```{r, results='hide'}
# RE-GRID LAYERS FASTER ----

layers_regrid <- regridR::regrid(layers = layers, grid = EEAgrid, 
                                 exactextract = TRUE)

terra::plot(layers_regrid)
```

We can visually check that the output (re-gridded) `layers`' pixels align with the input EEA `grid`:

```{r}
terra::plot(layers_regrid[[1]])

terra::plot(EEAgrid, lwd = 0.2, add = TRUE)
```

